---
title: US Healthcare Spending and Coverage
date: Sept 17, 2018
output: 
    html_document:
        theme: cosmo 
        toc: true
        toc_float: true
        highlight: tango
        number_sections: false
fig_width: 5
fig_height: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Motivation 

In the last lecture, we were analyzing health
care data in the United States. We were
exploring the following questions: 

1. Is there a relationship between healthcare coverage and healthcare spending in the United States? 
2. Which US states spend the most and which spend the least on healthcare? How does the spending distribution change across geographic regions in the United States?
3. How do healthcare coverage and spending relate to life expectancy? 

## Healthcare data
The data is from the [Henry J Kaiser Family Foundation (KFF)](https://www.kff.org). 

* [Health Insurance Coverage of the Total Population](https://www.kff.org/other/state-indicator/total-population/) - Includes years 2013-2016
* [Health Care Expenditures by State of Residence (in millions)](https://www.kff.org/other/state-indicator/health-care-expenditures-by-state-of-residence-in-millions/) - Includes years 1991-2014
* [Life Expectancy at Birth (in years)](https://www.kff.org/other/state-indicator/life-expectancy)

## Re-cap from last time

Let's recall what we did before: 

Load in the `tidyverse` of R package
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
```

Read in the datasets with the `readr` R package
```{r, message=FALSE, warning=FALSE}
coverage <- read_csv("../data/KFF/healthcare-coverage.csv", 
                     skip = 2, col_names = TRUE)
coverage <- read_csv("../data/KFF/healthcare-coverage.csv", 
                     skip = 2, col_names = TRUE,
                     n_max  = which(coverage$Location == "Notes")-1)
spending <- read_csv("../data/KFF/healthcare-spending.csv", 
                     skip = 2, col_names = TRUE)
spending <- read_csv("../data/KFF/healthcare-spending.csv", 
                     skip = 2, col_names = TRUE,
                     n_max  = which(spending$Location == "Notes")-1)
life <- read_csv("../data/KFF/life-expectancy.csv", 
                     skip = 2, col_names = TRUE)
life <- read_csv("../data/KFF/life-expectancy.csv", 
                     skip = 2, col_names = TRUE,
                     n_max  = which(life$Location == "Sources")-1)
```

Convert the `coverage` and `spending` datasets 
from _wide_ to _long_ formats using the `tidyr` 
R package
```{r, message=FALSE, warning=FALSE}
coverage <- gather(coverage, "year_type", "tot_coverage", -Location)
spending <- gather(spending, "year", "tot_spending", -Location)
```

Wrangle the datasets using some `dplyr` action verbs
```{r, message=FALSE, warning=FALSE}
coverage <- 
  coverage %>% 
  separate(year_type, sep="__", into=c("year", "type"), convert = TRUE) %>%
  mutate_at("tot_coverage", as.integer)

spending <- 
  spending %>% 
  separate(year, sep="__", into=c("year", "name"), convert = TRUE) %>% 
  select(-name)

life <- 
  life %>% 
  rename(life_exp_years = `Life Expectancy at Birth (years)`)
```

Now we will continue to learn more about data
wrangling using the `dplyr` package, including 
how to join two datasets together, how to summarize
columns of data, and how to apply group operations 
using the "split-apply-combine" concept. 

We will also start an introduction to the 
`ggplot2` package for data visualization. 

# Data Wrangling

Before we dive back into data wrangling with `dplyr`, 
it would be nice add some information about each state. 
Namely, the state abbreviation and state region 
(i.e. north, south, etc). 

For this we use the 
[state](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/state.html)
dataset in the `datasets` R package. 

Before we begin, let's look at what states are there: 

```{r}
unique(coverage$Location)
```

We see there are more than 50 states because 
"United States" and "District of Columbia" 
are both included. 

Let's look what states are inside the `state` dataset. 
```{r}
library(datasets)
data(state)
unique(state.name)
```

Ah, ok. So let's start by dealing with
DC as a special case. 

```{r}
state.abb <- c(state.abb, "DC")
state.region <- as.factor(c(as.character(state.region), "South"))
state.name <- c(state.name, "District of Columbia")
coverage$abb <- state.abb[match(coverage$Location, state.name)]
coverage$region <- state.region[match(coverage$Location, state.name)]

coverage
```

We will deal with the "United States" in
the next section. 

## The `dplyr` R package

### Join two datasets together

Up until now, we have been working with three
datasets `coverage`, `spending`, and `life` 
separtely. Next, we will combine these together. 

If we want to combine, say, `coverage` and `spending` 
together, we have to decide a few things. Both 
share a `Location` column and a `year` column. 
However, the range of `years` is different between
datasets. 

```{r}
table(coverage$year)
table(spending$year)
```

Do we want a dataset with all the years available
or only a portion of it? Because there is spending
information from 1991-2014, 
and coverage information from 2013-2016.

`dplyr` has a list of `join` functions that are 
useful to combine datasets. To read more about them, 
Jenny Bryan has created a nice 
[cheatsheet](http://stat545.com/bit001_dplyr-cheatsheet.html). 

If we look at the help file
```{r}
?dplyr::join
```

We see there are several options for us to pick from. 
Let's try one out. We'll start with `left_join()`
and see what that does. 

```{r}
hc <- left_join(coverage, spending, by = c("Location", "year"))
head(hc)
tail(hc)
```

What did it do? We see that the new `hc` dataset
includes all the years from 2013-2016 (as that 
is the range of years in `coverage`), but because
the `spending` dataset only goes to 2014, the 
`tot_spending` is reported as NA for years 2015 and
2016. 

What about a `right_join()`? 
```{r}
hc <- right_join(coverage, spending, by = c("Location", "year"))
head(hc)
tail(hc)
```

Here, we see every row in the spending dataset is 
there, but with NAs for the years that there was no 
coverage data. 

There is also a `full_join()` and 
`inner_join()`. If we want the intersection of 
`years` from `coverage` and `spending` (meaning only 
2013 and 2014), we should use `inner_join()`. 

```{r}
hc <- inner_join(coverage, spending, by = c("Location", "year"))
head(hc)
tail(hc)
```

Yes, that's what we want! 

Next, let's combine our new `hc` dataset 
with the `life` dataset.
```{r}
colnames(hc)
colnames(life)
```

In this case, 
the only column variable that
is shared between these two is the `Location` 
column. So we can use 

```{r}
hc <- left_join(hc, life, by = c("Location"))
hc
```

Finally, if we are only interested in looking at US 
states, we can remove the rows corresponding to 
the `Location == "United States"`

```{r}
hc <- hc %>% 
  filter(Location != "United States")
```

Another problem is that inside our `hc` 
dataset, we have seen there are 
multiple `types` of healthcare coverage

```{r}
table(hc$type)
```

The `total` type is not really a formal type of
healthcare coverage. It really represents just 
the total number of people in the state. This is 
useful information and we can include it as a 
column called `tot_pop`. How can we do this? 

Well, one way would be to use the `join` functions
again in `dplyr`. 

```{r}
pop <- hc %>% 
  filter(type == "Total") %>% 
  select(Location, year, tot_coverage)
pop

hc <- hc %>% 
  filter(type != "Total") %>% 
  left_join(pop, by = c("Location", "year")) %>% 
  rename(tot_coverage = tot_coverage.x, tot_pop = tot_coverage.y)
hc
```

We can check to make sure that the `total`
is no longer listed as a `type` of healthcare
coverage. 

```{r}
table(hc$type)
```

We are now ready to try answering our first 
question that we asked: 

> 1. Is there a relationship between healthcare coverage and healthcare spending in the United States?

Let's pick out the `type==Employer` 
and `year==2013`. 

```{r}
hc.employer.2013 <- hc %>%
  filter(type == "Employer", year == "2013")
plot(hc.employer.2013$tot_spending, 
     hc.employer.2013$tot_coverage, log = "xy", 
     xlab = "spending", ylab = "coverage")
```

We see there is a strong relationship. However, 
we also see that healthcare coverage and spending 
is also strongly related to population size 

```{r, fig.width=10, fig.height=4}
par(mfrow=c(1,2))
plot(hc.employer.2013$tot_pop, 
     hc.employer.2013$tot_coverage, log = "xy", 
     xlab = "population size", ylab = "coverage")
plot(hc.employer.2013$tot_pop, 
     hc.employer.2013$tot_spending, log = "xy", 
     xlab = "population size", ylab = "spending")
```

This means we need to take into account the 
population size of each state when we are 
comparing the heathcare coverage and spending. 

### Add columns using `mutate()`

Instead of the absolute number of people who 
are covered (`tot_coverage`), we will calculate
the proportion of people who are coverage in 
each state, year and type. 

For this, we will use the `mutate()` function 
in `dplyr`. 

```{r}
hc <- hc %>% 
    mutate(prop_coverage = tot_coverage/tot_pop) 
hc
```

We need to add another column to our dataset.
We will add the spending per capita (or spending 
per person) in dollars and name this column 
`spending_capita`. 

How we will do this?

The `tot_spending` column is reported 
in millions (1e6). Therefore, to calculate 
`spending_capita` we will need to adjust for this
scaling factor to report it on the original scale
(just dollars) and then divide by `tot_pop`.

```{r}
hc <- hc %>% 
    mutate(spending_capita = (tot_spending*1e6) / tot_pop) 
hc %>% select(prop_coverage, spending_capita)
```

Now we are ready to go back to our first question

> 1. Is there a relationship between healthcare coverage and healthcare spending in the United States? 

```{r}
hc.employer.2013 <- hc %>%
  filter(type == "Employer", year == "2013")
plot(hc.employer.2013$spending_capita, 
     hc.employer.2013$prop_coverage, log = "xy", 
     xlab = "spending per capita", 
     ylab = "proportion of Employer coverage")
```

Yes, it looks like there is a relationship for 
`Employer` healthcare coverage in 2013. 

We will continue to explore the other
types of coverages later on. For now, we 
get back to to learning more action verbs 
in `dplyr`. 

Our second question that we were interested 
in was: 

> 2. Which US states spend the most and which spend the least on healthcare? How does the spending distribution change across geographic regions in the United States?

To answer these questions, we need to learn how 
to calculate summary statistics in our data. 

### Create summaries of columns using `summarise()`

The `summarise()` function in `dplyr` 
will create summary statistics for a given 
column in the data frame 
such as finding the max, min, average. 
For example, to compute the average spending 
per capita, we can apply the `mean()` function 
to the column `spending_captia` and call the 
summary value `avg_spending_capita`. 

```{r}
hc %>% 
  summarise(avg_spending_capita = mean(spending_capita))
```

There are many other summary statistics you 
could consider such `sd()`, `min()`, `median()`, 
`mean()`, `sum()`, `n()` (returns the length of vector), 
`first()` (returns first value in vector), 
`last()` (returns last value in vector) and 
`n_distinct()` (number of distinct values in vector). 

Also note, this is the average across all states,
and all years. This is not very informative. 

If you recall, our question asked about 
_which states_ spent the most, so we want 
an average spending per capita for each state. 

For this, we need to introduc another function in 
`dplyr` called `group_by()`. 

### Group operations using `group_by()`

The `group_by()` verb is and incredibly powerful
function in `dplyr`. As we mentioned before
it's related to concept of "split-apply-combine". 

In our example above, we want to split the data 
frame by some variable (e.g. `Location`), 
apply a function to the individual 
data frames (`mean`) and then combine the output
back into a summary data frame. 

Let's see how that would look

```{r}
hc %>% 
  group_by(Location) %>%
  summarise(avg_spending_capita = mean(spending_capita))
```

That's better. Here we are averaging across the
years 2013 and 2014. 

#### Your turn 

What are the top 3 states that have the largest 
average spending per capita? What about the top 
3 states with the smallest average spending 
per capita? 

```{r}
## add your code here

```

How does the spending distribution change across 
geographic regions in the United States? 

**Hint**: Calculate the mean and standard deviation of
spending per capita for each geographic region in the US. 

```{r}
## your code here 

```

Another way to visualize distributions is to use boxplots. 

Create four boxplots represening the spending per capita 
distribution for each of the four regions using 
the `boxplot()` function in R.

```{r}
## add your code here

```

Now that we have our data in a `tidy` format, next, 
we will learn about how to do this using the 
`ggplot2` R package in the `tidyverse`. 

# Data Visualization

As you have already seen, there are many functions available
in base R that can create plots (e.g. `plot()`, `boxplot()`). 
Others include: `hist()`, `qqplot()`, etc. These 
functions are great because they come with a basic installation 
of R and can be quite powerful when you need a quick visualization 
of something when you are exploring data. 

We are choosing to introduce `ggplot2` because, in our 
opinion, it's one of the simplest ways for beginners to 
create relatively complicated plots that are intuitive 
and aesthically pleasing. 

## The `ggplot2` R package

The reasons [`ggplot2`](http://ggplot2.tidyverse.org) 
is generally intuitive for beginners is the use of 
[grammar of graphics](http://vita.had.co.nz/papers/layered-grammar.html) 
or the `gg` in `ggplot2`. The idea is that you can construct
many sentences by learning just a few nouns, adjectives,
and verbs. There are specific "words" that we will need to 
learn and once we do, you will be able to create 
(or "write") hundreds of different plots. 

The critical part to making graphics using `ggplot2` is the 
data needs to be in a _tidy_ format. Given that we have 
just spend the last two lectures learning about how to 
work with _tidy_ data, we are primed to take 
advantage of all that `ggplot2` has to offer! 

We will show how it's easy to pipe _tidy_ data
(output) as input to other functions that creates
plots. This all works because we are working 
within the _tidyverse_. 

## What is the `ggplot()` function? 

As explained by Hadley Wickham: 

> the grammar tells us that a statistical graphic is a mapping from data to aesthetic attributes (colour, shape, size) of geometric objects (points, lines, bars). The plot may also contain statistical transformations of the data and is drawn on a specific coordinates system.

### `ggplot2` Terminology 
* **ggplot** - the main function where you specify the data set and variables to plot (this is where we define the `x` and
`y` variable names)
* **geoms** - geometric objects
    * e.g. `geom_point()`, `geom_bar()`, `geom_line()`, `geom_histogram()`
* **aes** - aesthetics
    * shape, transparency, color, fill, linetype
* **scales** - define how your data will be plotted
    * continuous, discrete, log, etc

There are three ways to initialize a `ggplot()` object. 

An empty ggplot object
```{r}
p <- ggplot()         
```

A ggplot object associated with a dataset
```{r}
p <- hc %>% 
      filter(year==2014) %>% 
      ggplot()
```
or a ggplot object with a dataset and `x` and `y` defined

```{r}
p <- hc %>% 
      filter(year==2014) %>% 
      ggplot(aes(x = spending_capita, y = prop_coverage))
```

```{r, eval = FALSE}
p
```

### Creating your first `ggplot()`

The function `aes()` is an aesthetic mapping 
function inside the `ggplot()` object. We 
use this function to specify plot attributes 
(e.g. `x` and `y` variable names) that 
will not change as we add more layers.  

Anything that goes in the `ggplot()` object becomes
a global setting. From there, we use the `geom`
objects to add more layers to the base `ggplot()` 
object. These will define what we are interested in 
illustrating using the data.  

If you recall, our first question that we were
interested in was 

> 1. Is there a relationship between healthcare coverage and healthcare spending in the United States?

Before, we were using base R to create something
like this: 

```{r}
hc.employer.2013 <- hc %>% 
  filter(type == "Employer", year == "2013")

plot(hc.employer.2013$spending_capita, 
     hc.employer.2013$prop_coverage, 
     xlab = "spending per capita", 
     ylab = "coverage proportion")
```

Let's  re-create this plot with `ggplot2` 
using the `geom_point()` geometry. 

```{r}
p <- hc %>%
  filter(type == "Employer", year == "2013") %>% 
  ggplot(aes(x = spending_capita, y = prop_coverage)) 
p + geom_point() + 
  xlab("spending per capita") + 
  ylab("coverage proportion")
```

We used the `xlab()` and `ylab()` functions
in `ggplot2` to specify the x-axis and y-axis
labels. 

**Note**, we do not have to assign (`<-`) the plot 
to anything: 

```{r}
hc %>%
  filter(type == "Employer", year == "2013") %>% 
  ggplot(aes(x = spending_capita, y = prop_coverage)) + 
  geom_point() + 
  xlab("spending per capita") + 
  ylab("coverage proportion")
```


It's also simple to fit a linear regression model 
and plot it on top of scatter plot using the 
`geom_smooth()` (or `stat_smooth()`) functions. 

```{r}
hc %>%
  filter(type == "Employer", year == "2013") %>% 
  ggplot(aes(x = spending_capita, y = prop_coverage)) + 
  geom_point() + 
  xlab("spending per capita") + 
  ylab("coverage proportion") + 
  geom_smooth(method = "lm", col = "red")
```

The standard error bounds are computed and included 
in the plot. 

In the next lecture, we will explore more of `ggplot2`'s 
capabilities and (hopefully) finally answer the three 
questions we originally asked. 

### `ggplot2` cheatsheet

<img src="https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf" align="middle" width="700">




