---
title: US Healthcare Spending and Coverage
date: Sept 19, 2018
output: 
    html_document:
        theme: cosmo 
        toc: true
        toc_float: true
        highlight: tango
        number_sections: false
fig_width: 5
fig_height: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Motivation 

In the last two lectures, we were analyzing health
care data in the United States. We were
exploring the following questions: 

1. Is there a relationship between healthcare coverage and healthcare spending in the United States? 
2. Which US states spend the most and which spend the least on healthcare? How does the spending distribution change across geographic regions in the United States?
3. How do healthcare coverage and spending relate to life expectancy? 

The data is from the [Henry J Kaiser Family Foundation (KFF)](https://www.kff.org). 

* [Health Insurance Coverage of the Total Population](https://www.kff.org/other/state-indicator/total-population/) - Includes years 2013-2016
* [Health Care Expenditures by State of Residence (in millions)](https://www.kff.org/other/state-indicator/health-care-expenditures-by-state-of-residence-in-millions/) - Includes years 1991-2014
* [Life Expectancy at Birth (in years)](https://www.kff.org/other/state-indicator/life-expectancy)

Let's recall what we did before: 

Load in the `tidyverse` of R package
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
```

Read in the datasets with the `readr` R package
```{r, message=FALSE, warning=FALSE}
coverage <- read_csv("../data/KFF/healthcare-coverage.csv", 
                     skip = 2, col_names = TRUE)
coverage <- read_csv("../data/KFF/healthcare-coverage.csv", 
                     skip = 2, col_names = TRUE,
                     n_max  = which(coverage$Location == "Notes")-1)
spending <- read_csv("../data/KFF/healthcare-spending.csv", 
                     skip = 2, col_names = TRUE)
spending <- read_csv("../data/KFF/healthcare-spending.csv", 
                     skip = 2, col_names = TRUE,
                     n_max  = which(spending$Location == "Notes")-1)
life <- read_csv("../data/KFF/life-expectancy.csv", 
                     skip = 2, col_names = TRUE)
life <- read_csv("../data/KFF/life-expectancy.csv", 
                     skip = 2, col_names = TRUE,
                     n_max  = which(life$Location == "Sources")-1)
```

Convert the `coverage` and `spending` datasets 
from _wide_ to _long_ formats using the `tidyr` 
R package
```{r, message=FALSE, warning=FALSE}
coverage <- gather(coverage, "year_type", "tot_coverage", -Location)
spending <- gather(spending, "year", "tot_spending", -Location)
```

Wrangle the datasets using some `dplyr` action verbs
```{r, message=FALSE, warning=FALSE}
coverage <- 
  coverage %>% 
  separate(year_type, sep="__", into=c("year", "type"), convert = TRUE) %>%
  mutate_at("tot_coverage", as.integer)

spending <- 
  spending %>% 
  separate(year, sep="__", into=c("year", "name"), convert = TRUE) %>% 
  select(-name)

life <- 
  life %>% 
  rename(life_exp_years = `Life Expectancy at Birth (years)`)
```

Add abbreviation and region information to the 
`coverage` dataset

```{r}
library(datasets)
data(state)
unique(state.name)

state.abb <- c(state.abb, "DC")
state.region <- as.factor(c(as.character(state.region), "South"))
state.name <- c(state.name, "District of Columbia")
coverage$abb <- state.abb[match(coverage$Location, state.name)]
coverage$region <- state.region[match(coverage$Location, state.name)]
```


Join the datasets together
```{r, message=FALSE, warning=FALSE}
hc <- inner_join(coverage, spending, by = c("Location", "year"))
hc <- left_join(hc, life, by = c("Location"))
hc <- hc %>% filter(Location != "United States")
pop <- hc %>% 
  filter(type == "Total") %>% select(Location, year, tot_coverage)
hc <- hc %>% 
  filter(type != "Total") %>% 
  left_join(pop, by = c("Location", "year")) %>% 
  rename(tot_coverage = tot_coverage.x, tot_pop = tot_coverage.y)
```

Add created two new columns `prop_coverage` and 
`spending_capita`

```{r}
hc <- hc %>% 
    mutate(prop_coverage = tot_coverage/tot_pop,
           spending_capita = (tot_spending*1e6) / tot_pop) 
hc
```

Let's learn a little bit more about `ggplot2` to 
answer those questions using
the tools we have learned in the last two lectures 

# Data Visualization

## The `ggplot2` R package

In the previous lecture, we learned about 

1. the main `ggplot()` function
2. aesthic mappings or or _aes_ (connects the data we see on the graph; used as input to the geometry functions)
3. geometry functions or _geoms_ (adding layers on top of the `ggplot()` object)
4. scales or how the data will be plotted (e.g. continuous, discrete, log, etc)

### Creating scatter plots using `geom_point()`

We also created our first `ggplot()` object to explore 
the question 

> 1. Is there a relationship between healthcare coverage and healthcare spending in the United States? 

```{r}
hc %>%
  filter(type == "Employer", year == "2013") %>% 
  ggplot(aes(x = spending_capita, y = prop_coverage)) + 
  geom_point() + 
  xlab("spending per capita") + 
  ylab("coverage proportion") + 
  geom_smooth(method = "lm", col = "red")
```

It would be nice to know which state is represented
by which state. For this, we will introduce another 
_geom_ called `geom_text()`. 

### Adding layers of text using `geom_text()`

In our dataset, we have information about the 
abbreviation for each state. We could add the 
abbreviations for each state next to the point on 
the plot to assess which states have a higher or
lower coverage for a given amount of money they 
spend per capita. 

```{r}
hc %>% 
  filter(type == "Employer", year == "2013") %>% 
  ggplot(aes(x = spending_capita, y = prop_coverage)) + 
  geom_point() + 
  xlab("spending per capita") + 
  ylab("coverage proportion") + 
  geom_smooth(method = "lm", col = "red") + 
  geom_text(aes(label=abb))
```

That is cool, but it would be even better if we 
could _nudge_ the text over a bit. Let's look at 
the help file for `geom_text()`: 

```{r, eval=FALSE}
?ggplot2::geom_text
```

We see there is an argument called 
`nudge_x` and `nudge_y`. We can use these 
to _nudge_ the text over a bit so the 
text is not directly on top of the points. 

```{r}
hc %>% 
  filter(type == "Employer", year == "2013") %>% 
  ggplot(aes(x = spending_capita, y = prop_coverage)) + 
  geom_point() + 
  xlab("spending per capita") + 
  ylab("coverage proportion") + 
  geom_smooth(method = "lm", col = "red") + 
  geom_text(aes(label=abb), nudge_x = 150)
```

#### Your turn 

Color each point (or state) by what region 
they are from. 

```{r}
## add your code here

hc %>% 
  filter(type == "Employer", year == "2013") %>% 
  ggplot(aes(x = spending_capita, y = prop_coverage, 
             color = region)) + 
  geom_point() + 
  xlab("spending per capita") + 
  ylab("coverage proportion") + 
  geom_smooth(method = "lm", col = "red") + 
  geom_text(aes(label=abb), nudge_x = 150)
```


### faceting across a variable

Ok, getting back to our original question: 

> 1. Is there a relationship between healthcare coverage and healthcare spending in the United States? 

We saw there was a positive relationship,
but this was only for one type of healthcare 
coverage (`Employer`) and one year. 
What about the other types? 

For this, we will introduce `facets`. The idea 
of _faceting_ is to stratify the data by some 
variable and make the same plot for each strata. 

For example, if we wanted to _facet_ by the 
`type` variable, we will add a layer to our 
`ggplot()` object using the `facet_grid()` or 
`facet_wrap()` functions. The function expects
the row and column variables to be separated 
by a `~`. 

```{r, fig.width=12, fig.height=8}
hc %>%
  filter(year == "2013") %>% 
  ggplot(aes(x = spending_capita, y = prop_coverage, 
             color = region)) + 
  geom_point() + 
  xlab("spending per capita") + 
  ylab("coverage proportion") + 
  geom_smooth(method = "lm", col = "red") + 
  geom_text(aes(label=abb), nudge_x = 150) + 
  facet_wrap(~type)
```

We see that the proportion of people covered 
have different scales in the y-axis. Let's 
read the help file to see if there is some way 
to not restrict the y-axis to be the same. 
```{r}
?ggplot2::facet_grid
```

Yes, we see there is an argument called 
`scales` that can be `free_y`, (free columns),
`free_x` (free rows), and `free` (both). 
Let's try `free_y` and look at a different
year (`year=="2014"`): 

```{r, fig.width=12, fig.height=6}
hc %>%
  filter(year == "2014") %>% 
  ggplot(aes(x = spending_capita, y = prop_coverage, 
             color = region)) + 
  geom_point() + 
  xlab("spending per capita") + 
  ylab("coverage proportion") + 
  geom_smooth(method = "lm", col = "red") + 
  geom_text(aes(label=abb), nudge_x = 150) + 
  facet_wrap(~type, scales="free_y")
```

Given we know `Other Public` refers to the 
military or Veterans Adminstration, we can see 
states like HI, VA, NV have a larger 
proportion of military or VA `Other Public` 
type coverage. While a state like AK has a 
similar proportion of `Other Public` coverage, 
it has a much larger spending per capita. 

We also see a negative relationship with the 
`Uninsured` type. The more states spend, the
less uninsured people in the state. 

### Creating boxplots using `geom_boxplot()`

Next, let's revisit the second question. 

> 2. Which US states spend the most and which spend the least on healthcare? How does the spending distribution change across geographic regions in the United States?

Let's try making a boxplot with `ggplot2`. 
If you recall, the way to do this in base R was: 

```{r}
boxplot(hc$spending_capita ~ hc$region)
```

Now, we introduce the `geom_boxplot()` 
function. Note, we needed to tell `ggplot2`
what needs to be along the x and y axis in
`aes()`. 

```{r}
hc %>% 
  ggplot(aes(x = region, y = spending_capita)) + 
  geom_boxplot()
```

### faceting by two variables 

> 3. How do healthcare coverage and spending relate to life expectancy? 

In our last section, let's first explore 
the relationship between spending and 
life expectancy using facets. We were 
faceting by healthcare `type`, so let's 
stick with that, but change the y-axis
to `life_exp_years`. 

```{r, fig.width=12, fig.height=8}
p <- hc %>%
  filter(year == "2014") %>% 
  ggplot(aes(x = spending_capita, y = life_exp_years, 
             color = region)) + 
  geom_point() + 
  xlab("spending per capita") + 
  ylab("life expectactancy") + 
  geom_smooth(method = "lm", col = "red") + 
  geom_text(aes(label=abb))

p + facet_wrap(~type, scales="free")
```

Hmm... So looks like there is a positive 
association for all the types (more you spend
the longer the life expectacy). But there is 
something strange about the fact that the regions
have very different life expectancies. 

Let's try faceting by both `region` and `type`. 
Note that we can facet by rows putting a 
column name before the `~` and facet by 
columns putting a column name after the `~`. 
We are also using `facet_grid()` instead of 
`facet_wrap()`.

```{r, fig.width=12, fig.height=8}
p + facet_grid(region~type, scales="free")
```

What trends do you see? 

Finally, let's do the same for the 
`prop_coverage` column instead of 
`spending_capita` column to explore the 
relationship between life expectancy 

```{r, fig.width=12, fig.height=8}
p <- hc %>%
  filter(year == "2014") %>% 
  ggplot(aes(x = prop_coverage, y = life_exp_years, 
             color = region)) + 
  geom_point() + 
  xlab("coverage proportion") + 
  ylab("life expectactancy") + 
  geom_smooth(method = "lm", col = "red") + 
  geom_text(aes(label=abb))

p + facet_grid(region~type, scales="free")
```

What trends do you see? How would you summarize
these results? What's most interesting?

### `ggplot2` cheatsheet

<img src="https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf" align="middle" width="700">

# Mini-Case Study 1

For the last 20 mins of class, we will break 
into groups and work on a mini-case study. 
You can find the case study on the 
[course website](https://jhu-advdatasci.github.io/2018/). 


