---
title: "Dates and Times in R"
date: October 3, 2018
output: 
    html_document:
        theme: cosmo 
        toc: true
        toc_float: true
        highlight: tango
        number_sections: false
fig_width: 5
fig_height: 5
---

```{r, include=FALSE}
options(tibble.print_min = 5)
knitr::opts_chunk$set(comment = NA, collapse = FALSE)
```


R has developed a special representation of dates and times

- Dates are represented by the `Date` class

- Times are represented by the `POSIXct` or the `POSIXlt` class

- Dates are stored internally as the number of days since 1970-01-01

- Times are stored internally as the number of seconds since
  1970-01-01

# The `lubridate` package

* The `lubridate` package is a very useful package for dealing with all the little annoying aspects of dates/times

* Largely replaces the default date/time functions in base R

* Methods for date/time arithmetic

* Handles time zones, leap year, leap seconds, etc.

```{r,eval=FALSE}
install.packages("lubridate")
## Not part of `tidyverse` package
```

# Dates in R

Dates are represented by the `Date` class and can be coerced from a character string using the `ymd()` function.

```{r,message=FALSE}
library(lubridate)
x <- ymd("1970-01-01")
x
class(x)
unclass(x)

x <- ymd("2019-10-03")
unclass(x)
```

`Date` objects have their own special print methods that will always format as "YYYY-MM-DD".

## Alternate Formulations

Different locales have different ways formatting dates

```{r}
ymd("2016-09-13")  ## International standard
ymd("2016/09/13")  ## Just figure it out
mdy("09-13-2016")  ## Mostly U.S.
dmy("13-09-2016")  ## Europe
```
All of the above are valid and lead to the exact same object.


Even if the individual dates are formatted differently, `ymd()` can usually figure it out.

```{r}
x <- c("2016-04-05", 
       "2016/05/06",
       "2016,10,4")
ymd(x)
```

# Times in R

Times are represented using the `POSIXct` or the `POSIXlt` class


- `POSIXct` is just a very large integer under the hood; it is a useful class when you want to store times in something like a data frame

- `POSIXlt` is a list underneath and it stores a bunch of other useful information like the day of the week, day of the year, month, day of the month

Times are represented as the number of *seconds* since 1970-01-01 00:00:00.

```{r}
x <- ymd_hms("2019-10-03 13:30:00")
class(x)
unclass(x)
```


If you want to know more about the international date/time standard, you can read about [ISO Standard 8601](https://en.wikipedia.org/wiki/ISO_8601).

## Inputing Time Data

Times can be coerced from a character string with `ymd_hms()`

```{r}
ymd_hms("2016-09-13 14:00:00")
ymd_hms("2016-09-13 14:00:00", tz = "America/New_York")
ymd_hms("2016-09-13 14:00:00", tz = "")
```

# Time Zones!

Time zones were created to make your data analyses more difficult.

* `ymd_hms()` function will by default use UTC as the time zone

* Specifying `tz = ""` will use the local time zone

* Better to specify time zone when possible to avoid ambiguity

You can go to Wikipedia to find the [list of time zones](https://en.wikipedia.org/wiki/List_of_tz_database_time_zones)

* Daylight savings time

* Some states are in two time zones

* Southern hemisphere is opposite


# Specifying Times in R

Finally, there is the `strptime()` function in case your dates are
written in a different format

```{r,tidy=FALSE}
datestring <- c("January 10, 2012 10:40", 
                "December 9, 2011 9:10")
x <- strptime(datestring, "%B %d, %Y %H:%M", 
              tz = "America/Los_Angeles")
x
```

* Check `?strptime` for details of formatting strings

* When reading in data with `read_csv()`, you may need to read in as character first and then convert to date/time


# Operations on Dates and Times

## Arithmetic 

You can add and subtract dates and times. You can do comparisons too (i.e. `==`, `<=`)

```{r, error=TRUE}
x <- ymd("2012-01-01", tz = "")  ## Midnight
y <- dmy_hms("9 Jan 2011 11:34:21", tz = "")
x - y
x + y  ## Nope!
```



Add a second to a time
```{r}
y + 1
```
Just keep the date portion
```{r}
y <- date(y)
y
```
Add a number to the date (in this case 1 day)
```{r}
y + 1  
```

## Leaps and Bounds

Even keeps track of leap years, leap seconds, daylight savings, and time zones.

Leap years
```{r}
x <- ymd("2012-03-01")
y <- ymd("2012-02-28")
x - y
```


Beware of time zones!
```{r}
x <- ymd_hms("2012-10-25 01:00:00", tz = "")
y <- ymd_hms("2012-10-25 06:00:00", tz = "GMT")
y - x
```

There are also *leap seconds*.

```{r}
.leap.seconds
```

# Extracting Elements of Dates/Times

There are a set of helper functions in `lubridate` that can extract sub-elements of dates/times

## Date Elements
```{r}
x <- ymd_hms(c("2012-10-25 01:13:46",
               "2015-04-23 15:11:23"), tz = "")
year(x)
month(x)
day(x)
weekdays(x)
```

## Time Elements

```{r}
x <- ymd_hms(c("2012-10-25 01:13:46",
               "2015-04-23 15:11:23"), tz = "")
minute(x)
second(x)
hour(x)
week(x)
```



# Visualizing Dates


## Reading in the Data

```{r, warning=FALSE, message=FALSE}
library(readr)
storm <- read_csv("../data/storm_events_2002.csv.gz", progress = FALSE)
names(storm)
```

Let's take a look at the BEGIN_DATE_TIME and DEATHS_DIRECT variables

```{r,message=FALSE}
library(dplyr)
select(storm, BEGIN_DATE_TIME, EVENT_TYPE, DEATHS_DIRECT)
```

We can first convert the date/time to a date/time R object.

```{r}
storm_sub <- select(storm, BEGIN_DATE_TIME, EVENT_TYPE, DEATHS_DIRECT) %>%
  mutate(begin = dmy_hms(BEGIN_DATE_TIME)) %>%
  rename(type = EVENT_TYPE,
         deaths = DEATHS_DIRECT) %>%
  select(begin, type, deaths)
storm_sub
```


## Histograms of Dates/Times


We can make a histogram of the dates/times to get a sense of when storm events occur.

```{r}
library(ggplot2)
storm_sub %>%
  ggplot(aes(x = begin)) + 
  geom_histogram(bins = 20) + 
  theme_bw()
```

We can group by event type too.

```{r,fig.width=12,fig.height=12}
library(ggplot2)
storm_sub %>%
  ggplot(aes(x = begin)) + 
  facet_wrap(~ type) + 
  geom_histogram(bins = 20) + 
  theme_bw() + 
  theme(axis.text.x.bottom = element_text(angle = 90))
```

## Scatterplots of Dates/Times

```{r}
storm_sub %>%
  ggplot(aes(begin, deaths)) + 
  geom_point()
```

If we focus on a single month, the x-axis adapts.

```{r}
storm_sub %>%
  filter(month(begin) == 6) %>%
  ggplot(aes(begin, deaths)) + 
  geom_point()

```

Similarly, we can focus on a single day.

```{r}
storm_sub %>%
  filter(month(begin) == 6, day(begin) == 16) %>%
  ggplot(aes(begin, deaths)) + 
  geom_point()

```

# Summary

- Dates and times have special classes in R that allow for numerical and statistical calculations

- Dates use the `Date` class

- Times use the `POSIXct` and `POSIXlt` class

- Character strings can be coerced to Date/Time classes using the `ymd()` and `ymd_hms()` functions. In strange cases, you can use the `strptime()`or the `as.Date()` functions.

- The `lubridate` package is essential for manipulating date/time data

- Both `plot` and `ggplot` "know" about dates and times and will handle axis labels appropriately

