---
title: "Regular Expressions in R"
date: October 8, 2018
output: 
    html_document:
        theme: cosmo 
        toc: true
        toc_float: true
        highlight: tango
        number_sections: false
fig_width: 5
fig_height: 5
---

```{r, include=FALSE}
options(tibble.print_min = 5)
knitr::opts_chunk$set(comment = NA, collapse = FALSE)
```


## What are Regular Expressions?

-   Regular expressions can be thought of as a combination of literals
    and *metacharacters*

-   To draw an analogy with natural language, think of literal text
    forming the words of this language, and the metacharacters defining
    its grammar

-   Regular expressions have a rich set of metacharacters


## Primary Metacharacters

The world of regular expressions is large, but there are few key metacharacters that get used often.

* Literals

* Beginning/End of line

* Character classes

* Repetition

* Parenthesized subexpression


### Literals

Literals are just sequences of characters. They have no special meaning and are interpreted by R *literally*.

```{r,echo=FALSE}
library(readr)
book <- read_lines("../data/book.txt")
```


The word "brother" is a literal

```{r,echo=FALSE,message=FALSE}
set.seed(3)
library(dplyr)
grep("brother", book, value = TRUE) %>%
  sample(5) %>%
  writeLines(sep = "\n\n")
```

So is "Brother" (but not the same!).

```{r,echo=FALSE,message=FALSE}
set.seed(4)
grep("Brother", book, value = TRUE, perl = TRUE) %>%
  writeLines(sep = "\n\n")
```

### Beginning/End of Line

* `^` indicates match at the **beginning** of a line
* `$` indicates match at the **end** of a line

`^The`

```{r,echo=FALSE}
set.seed(44)
grep("^The", book, value = TRUE) %>%
  sample(5) %>%
  writeLines(sep = "\n\n")
```

`the$`

```{r,echo=FALSE}
set.seed(44)
grep("the$", book, value = TRUE) %>%
  sample(5) %>%
  writeLines(sep = "\n\n")
```


### Character Classes

The square brackets `[` and `]` indicate classes of characters to search for.

`[a-z]`
```{r,echo=FALSE}
set.seed(44)
grep("[a-z]", book, value = TRUE) %>%
  sample(3) %>%
  writeLines(sep = "\n\n")
```

`[A-Z]`
```{r,echo=FALSE}
set.seed(44)
grep("[A-Z]", book, value = TRUE) %>%
  sample(3) %>%
  writeLines(sep = "\n\n")
```

`[0-9]`
```{r,echo=FALSE}
set.seed(44)
grep("[0-9]", book, value = TRUE) %>%
  sample(3) %>%
  writeLines(sep = "\n\n")
```

`[Tt]he `
```{r,echo=FALSE}
set.seed(44)
grep("[Tt]he", book, value = TRUE) %>%
  sample(3) %>%
  writeLines(sep = "\n\n")
```

You can now combine character classes with begninning/end of line markers.

`^[Tt]he`
```{r,echo=FALSE}
set.seed(44)
grep("^[Tt]he", book, value = TRUE) %>%
  sample(3) %>%
  writeLines(sep = "\n\n")
```


### Matching Anything

The `.` is used to match anything, including nothing.

`9.1`

```{r,echo=FALSE}
set.seed(44)
grep("9.1", book, value = TRUE) %>%
  sample(3) %>%
  writeLines(sep = "\n\n")
```



### Repetition


* The `+` is used to indicate "repeat the immediately preceding symbol 1 or more times"

* The `*` is used to indicate "repeat the immediately preceding symbol 0 or more times"

* `{}` can be used to indicate a range of repetition


`[0-9]+`

```{r,echo=FALSE}
set.seed(4090)
grep("[0-9]+", book, value = TRUE) %>%
  sample(3) %>%
  writeLines(sep = "\n\n")
```


`2[0-9]*`

```{r,echo=FALSE}
set.seed(40)
grep("2[0-9]*", book, value = TRUE) %>%
  sample(3) %>%
  writeLines(sep = "\n\n")
```


`[0-9]{4,6}`

```{r,echo=FALSE}
set.seed(4090)
grep("[0-9]{4,6}", book, value = TRUE) %>%
  sample(3) %>%
  writeLines(sep = "\n\n")
```


` he .* good`

```{r,echo=FALSE}
set.seed(4090)
grep(" he .* good", book, value = TRUE) %>%
  sample(3) %>%
  writeLines(sep = "\n\n")
```


### Parenthesized Subexpression

* `()` can be used to "capture" subexpressions


`([0-9]+)\1`

```{r,echo=FALSE}
set.seed(4090)
grep("([0-9]+)\\1", book, value = TRUE) %>%
  sample(3) %>%
  writeLines(sep = "\n\n")
```


`\((.*)\)`

```{r,echo=FALSE}
set.seed(4090)
grep("\\((.*)\\)", book, value = TRUE) %>%
  sample(3) %>%
  writeLines(sep = "\n\n")
```




## Regular Expression Functions

The primary R functions for dealing with regular expressions are

* `grep`, `grepl`: Search for matches of a regular
  expression/pattern in a character vector; either return the indices
  into the character vector that match, the strings that happen to
  match, or a TRUE/FALSE vector indicating which elements match

* `regexpr`, `gregexpr: Search a character vector for regular
  expression matches and return the indices of the string where the
  match begins and the length of the match

* `sub`, `gsub`: Search a character vector for regular
  expression matches and replace that match with another string

* `regexec`: Easier to explain through demonstration.

### grep

```{r loaddata,cache=TRUE}
library(readr)
commits <- read_lines("../data/commit_logs_strip.txt.bz2")
head(commits)
```


How many commits are there?

```{r grep1,cache=TRUE}
g <- grep("^commit", commits)
head(g)
length(g)
```


Sometimes you want `grep()` to return the value instead of the index

```{r grep2,cache=TRUE}
g <- grep("^commit", commits, value = TRUE)
head(g)
```


Who are the authors of these commits?

```{r,cache=TRUE,warning=FALSE}
g <- grep("^Author", commits, value = TRUE, perl = TRUE)
head(g)
length(unique(g))
```


### grepl

By default, `grep()` returns the *indices* into the character vector where the regex pattern matches.

```{r}
head(state.name)
grep("^New", state.name)
```

`grepl()` returns a logical vector indicating which element matches.

```{r}
i <- grepl("^New", state.name)
head(i, 10)
```


Some limitations of `grep()`:

* The `grep()` function tells you which strings in a character vector match a certain pattern but it doesn’t tell you exactly where the match occurs or what the match is (for a more complicated regex).

* The `regexpr()` function gives you the index into each string where the match begins and the length of the match for that string.

* `regexpr()` only gives you the first match of the string (reading left to right). `gregexpr()` will give you all of the matches in a given string.

### regexpr

How can we obtain the email addresses of the authors?

```{r,cache=TRUE}
commits[12:24]
```

What if we use the regex `<(.*)>` and search for that?


We need to search the Author line for a pattern. We can first `grep` the Author lines and then search for a pattern.

```{r,cache=TRUE}
author <- grep("^Author:", commits, value = TRUE)
head(author, 3)
r <- regexpr("<.*>", author)
str(r)
```


* `regexpr()` returns a vector of integers indicating where the match starts

* The attribute `match.length` indicates how long the match is

* If there's no match, `regexpr()` returns `-1` with a `match.length` of `-1`.

The obvious way to select out a match is to use the indices and the `substr()` function.

```{r}
substr(author[1], 15, 15 + 6 - 1)
substr(author[3], 14, 14 + 6 - 1)
```


### regmatches

We can also use the `regmatches()` function to just grab all of the matches at once.

```{r,cache=TRUE}
r <- regexpr("<.*>", author)
m <- regmatches(author, r)
head(m)
```


### sub/gsub

But we still don't have actual email addresses. We need to remove the `<` and `>` characters. We can use the `sub()` function for that.

```{r}
sub("<", "", m[1:5])
sub(">", "", m[1:5])
```
But we want to remove *both* `<` and `>`!


We can use a regular expression in `sub()`.

```{r}
sub("<|>", "", m[1:5])
```

`gsub()` substitutes all occurrences of the regex (`g` is for "global").

```{r}
gsub("<|>", "", m[1:5])
```


### regexec

The `regexec()` function can make the previous task a bit simpler by using *parenthesized sub-expressions*.

```{r}
author[1]
```

We can capture the email address portion of the line with parentheses.

```{r}
regexec("^Author: [^ ]+ <(.*)>", author[1])
```


```{r}
r <- regexec("^Author: [^ ]+ <(.*)>", author[1])
regmatches(author[1], r)
```


When were all of the commits made?

```{r,cache=TRUE,warning=FALSE}
r <- regexec("^Date: +(.*)$", commits, perl = TRUE)
m <- regmatches(commits, r)
```


Now we can subset the elements that match

```{r,cache=TRUE}
library(purrr)
u <- map_int(m, length) > 0
head(m[u])
```

Finally, we can parse the dates/times.

```{r,cache=TRUE}
library(lubridate)
dates <- map_chr(m[u], 2) %>%
  parse_date_time("abd HMS Y z", tz = "America/New_York",
                  quiet = TRUE)
```

### Histogram

You can make a histogram of the dates

```{r}
hist(dates, "month", freq = TRUE)
rug(dates)
```

## Summary

The primary R functions for dealing with regular expressions are

* `grep`, `grepl`: Search for matches of a regular expression/pattern in a character vector

* `regexpr`, `gregexpr`: Search a character vector for regular expression matches and return the indices where the match begins; useful in conjunction with regmatches

* `sub`, `gsub`: Search a character vector for regular expression matches and replace that match with another string

* `regexec`: Gives you indices of parethensized sub-expressions.

















