---
title: 'Homework 2 (712): Predicting a diagnosis of malignant mesotheliomas'
output:
  html_document: default
  pdf_document: default
---

```{r, include = FALSE}
knitr::opts_chunk$set(include = TRUE, comment = NA)
```


## Preface 

* **This homework is due Monday November 23, 2018 at 11:59 PM**. 
* When you have completed the assignment, knit the R Markdown, commit your changes and push to GitHub.
* If you do not include axis labels and plot titles, then points will be deducted.
* If you do not include prose/text after the sections titled "Add a summary of your findings here", then points will be deducted. 
* As as reminder, you can use up to two late days on this assignment (if you have them available) without any penalty (see [Syllabus](https://jhu-advdatasci.github.io/2018/syllabus.html) on course website for more details on Late Day Policy).
* You are welcome and encouraged to discuss homework problems with others in order to better understand it, but the work you turn in must be your own. You must write your own code, data analyses, and communicate and explain the results in your own words and with your own visualizations. **All students turning in plagiarized solutions will be reported to Office of Academic Integrity, and will fail the assignment**.


## Motivation

Malignant mesotheliomas (MM) are very aggressive tumors. 
While there is an association to asbestos exposure, it may 
also be related to previous simian virus 40 (SV40) infection 
and quite possible for genetic predisposition. Molecular 
mechanisms and rural living are also associated with the
development of mesothelioma. 


### Overall Objective

The goal of this assignment is to develop a model to predict
the diagnosis of MM. Early detection of mesotheliomas helps
physicians to design a personalized treatment for saving lives.
You will explore the data and practice building classification 
models. You will also explore the limitations and uncertainties
of various approaches. During the course of your analysis there
will be many options for you to explore and various approaches
that you may take. Part of your job will be to choose amongst 
these many options and focus on a specific approach that you 
find most interesting or has the greatest potential for success.

### Data 

The data come from the 
[UCI Machine Learning](https://archive.ics.uci.edu/ml/datasets/Mesothelioma’s+disease+data+set+) 
repository. As stated on the website: 

> The mesotheliomas disease data set were prepared at Dicle 
University Faculty of Medicine in Turkey. Three hundred and 
twenty-four Mesothelioma patient data. In the dataset, all 
samples have 34 features.

You can access the data [here](https://archive.ics.uci.edu/ml/machine-learning-databases/00351/) 
and read about the 34 features on the 
[UCI Machine Learning](https://archive.ics.uci.edu/ml/datasets/Mesothelioma’s+disease+data+set+) 
page. 

# Problem 1: Exploring the Mesotheliomas data set

## Problem 1.1

Write code to test if the dataset (`Mesothelioma data set.xlsx`)
exists locally on your computer. If not, then download the file. 
Read in the `Mesothelioma data set.xlsx` in the `Data` sheet
using the `readxl` package. 

**Note**: Check out the `Attribute` and `Description` sheets
in the `.xlsx` file. 

```{r}
## add your code here 

```


## Problem 1.2

Use your exploratory data analysis skills to explore
the features in the dataset. For example, create summary
statistics, tables and/or plots to better understand the 
features. Summarize your findings.

```{r}
## add your code here 

```

#### Add a summary of your findings here



## Problem 1.3

Use your exploratory data analysis skills to explore 
the relationship between whether or not the patient was
diagnosed with MM and how that differs across age and gender? Are
there any interesting features related to age or gender?
Summarize your findings.

```{r}
## add your code here 

```

#### Add a summary of your findings here



## Problem 1.4

Use your exploratory data analysis skills to explore 
if there any other interesting features that contribute
to the diagnosis. Summarize your findings.

```{r}
## add your code here 

```

#### Add a summary of your findings here


## Problem 1.5

Split the data into a `train_set` (70%
of the data) and `test_set` (30% of the data). 

```{r}
## add your code here 

```



# Problem 2: Model building

In this problem, you will explore different approaches
to predict the diagnosis of MM. In each case, use 
10-fold cross validation to build a model using the 
`train_set`. 

## Problem 2.1: GLM approach

Fit a logistic regression model to predict the 
diagnosis of MM. Which variables did you include 
in your model? Summarize your findings.

**Things you may want to consider**: 

* It looks like the variable `diagnosis method` perfectly matches the `class of diagnosis`, so we may want to exclude this variable in the following analysis.)
* Model selection. Consider different approaches for model selection. For example, check out the `step()` function in the `stats` R package. What approach did you chooose and why? 

```{r}
## add your code here 

```

#### Add a summary of your findings here


## Problem 2.2

Train and tune a support vector machine to predict
the diagnosis of MM. Summarize your findings.

```{r}
## add your code here 

```

#### Add a summary of your findings here


## Problem 2.3

Tune and fit a random forest modelto predict
the diagnosis of MM. Summarize your findings.
 
```{r}
## add your code here 

```


#### Add a summary of your findings here


## Problem 2.4 

Predict the diagnosis of MM using the `test_set`
for each of the three approaches. Explore 
performance metrics and plots for each of the 
three approaches. Summarize your findings. 

```{r}
## add your code here 

```


#### Add a summary of your findings here




## Problem 3: Narrative

Given the results you produced in Problems 1 and 2,
it's time to summarize your results and create a summary 
of which method performed best and the choices you made 
in model fitting and tuning.

To get full credit for the problem, you will need the following: 

* Write a one paragraph abstract summarizing your findings 
from Problems 1-2 above. 

* Produce one summary figure/table/graphic/etc.
that provides evidence which approach worked best. 

* Produce a figure/table/graphic/etc. that describes 
other possible features that would be interesting to explore
that would be useful in deciding predicting the diagnosis 
of MM.

* Write one paragraph outlining the limitations of your 
analysis and what could be done in the future to address
those limitations.


