---
title: "Homework 3: University Academic Scorecard"
output: html_document
---

#### Preface 

* **This homework is due Friday October 26, 2018 at 11:59 PM**. 
* When you have completed the assignment, knit the R Markdown, commit your changes and push to GitHub.
* If you do not include axis labels and plot titles, then points will be deducted.
* If you do not include prose/text after the sections titled "Add a summary of your findings here", then points will be deducted. 
* As as reminder, you can use up to two late days (if available) on this assignment without any penalty (see [Syllabus](https://jhu-advdatasci.github.io/2018/syllabus.html) on course website for more details on Late Day Policy).
* You are welcome and encouraged to discuss homework problems with others in order to better understand it, but the work you turn in must be your own. You must write your own code, data analyses, and communicate and explain the results in your own words and with your own visualizations. **All students turning in plagiarized solutions will be reported to Office of Academic Integrity, and will fail the assignment**.

# Motivation 

You are a high school student applying to colleges
and you think you are interested in getting a Bachelor's 
in a health related field. 

Upon some googling, you are quickly overwhelmed 
at the sheer number of health related programs that exist.
Also, your parents have graciously offered to pay for
four years of college, but need you stay within a $50,000
budget (including everything e.g. tuition, fees, room and 
board, books, etc). 

Your parents told you that in 2013, President Obama 
announced that his administration would create a 
"[college scorecard](https://collegescorecard.ed.gov)
that parents and students can use to compare schools 
based on a simple criteria-where you an get the most
bang for your educational buck." 
[link to quote](https://www.brookings.edu/opinions/understanding-the-college-scorecard/). 

Your goal in this homework assignment is to explore the 
landscape of colleges (various criteria and cost) that 
have health related programs and identify a group of colleges 
that you would interested in applying to while staying 
within your budget. 

## Data 

The data that we will use are:

* [College Scorecard Data](https://collegescorecard.ed.gov/data/): The College Scorecard Database records summary information of over 7000 colleges and universities in details, which includes location, size, admission SAT score, degrees, graduation rate and so on. Here, instead of using the full data, we explore the most recent data, which can be [downloaded here](https://ed-public-download.app.cloud.gov/downloads/Most-Recent-Cohorts-All-Data-Elements.csv). (Notice that the data set is 132MB.) 
* Explanations of the column names can be found [here](https://collegescorecard.ed.gov/assets/CollegeScorecardDataDictionary.xlsx).
* Your are welcome to incorporate other data into your analysis, if necessary, as long as those other data sources are documented and their origin is clear.


## Problem 1: Exploring the Data

Read in the the College Scorecard Data and 
call it `univ_scorecard`. Filter out 
universities with a highest degree lower 
than Bachelor's degree and remove them (because 
we want to get a Bachelor degree). Then, using 
the columns containing information about the 
average cost of attendance (academic year 
institutions), use exploratoray data analysis 
(i.e. must include a plot here)
tools to see if there a difference in 
distributions of cost among public institutes, 
private nonprofit institutes and private 
for-profit institutes? 

Based on your budget (up to $50,000 a year), 
filter out universities outside of your budget. 
How many universities remain in your search scope?

**Hint**: 

1. You can refer to the data dictionary 
([download here](https://collegescorecard.ed.gov/assets/CollegeScorecardDataDictionary.xlsx)) 
for explanations of column names. 
2. Missing values can be omitted.

```{r}
## add your code here

```


## Problem 2: Identifying universities with health related degrees

### Problem 2.1 

As you are mostly considering getting a degree in 
a health related field, you want to identify 
groups of colleges that have a high percentage of
degrees awarded for health related fields. Read 
about the data columns from the data dictionary 
that are related to percentage of degrees awarded. 

In which field does the Johns Hopkins University have 
the most degrees awarded? What is the percentage? 
Is the Johns Hopkins University in your budget?

```{r}
## add your code here

```

#### Add a summary of your findings here


### Problem 2.2

For universities within our search scope (defined 
in Problem 1) and using the data about the 
percentage of degrees award in different fields, 
use dimensionality reduction methods and 
apply unsupervised methods (e.g. clustering approaches)
to explore and identify groups of universities 
that you might be interested in applying to. Are 
there groups of universties that cluster together? 
If so, how many groups? What do they cluster by?

For example, you could start about by using PCA 
and plot the universities along the first two PCs and
color them by percent of degrees awarded in 
"Health Professions and Related Programs". Are the
loadings useful? 

To get full credit, a complete response 
should include: (1) the application of at least
one dimensionality reduction method to explore
the most informative features in this data, (2) the 
application of at least one unsupervised analysis
(e.g. k-means) to identify groups of universities, 
(3) data visualizations or summary statistics, and
(4) a summary of your findings. 

```{r}
## add your code here

```

#### Add a summary of your findings here


## Problem 3: What other features are most important to you? 

In the data set, there are many other features that 
may help you pick a university, for example,

* admission rate and average SAT scores
* completion rate
* earnings after graduation
* geography
* size of university
* type: e.g. public or private, single sex or coed, research vs teaching focused, religious affiliation
* ....

Explore other features that are most important to you
in making a decision for colleges. 

```{r}
## add your code here

```

**Note**: There is no wrong answer. To get full credit, a 
complete response should include: (1) the features you 
are most interested in when considering universities,
(2) the approach you took to identify 
your topic choices, (3) data visualizations 
or summary statistics, and (4) a summary of your 
findings for this problem. 

**Note**: You may also use data outside of the scorecard,
as long as those other data sources are documented and 
their origin is clear.


## Problem 4: Narrative

Given the results you produced in Problems 1-3,
it's time to summarize your results and make a set of 
recommendations for which colleges you want to apply 
to with a health related program keeping in mind your 
budget. 

To get full credit for the problem, you will need the following: 

* Write a one paragraph abstract summarizing your findings 
from Problems 1-3 above. The paragraph should include 
a set of recommendations for your top three choices of 
colleges to apply to. 

* Produce one summary figure/table/graphic/etc.
that provides evidence supporting the recommendations made 
in the abstract above.

* Produce a figure/table/graphic/etc. that describes 
other possible features that would be interesting to explore
that would be useful in deciding which colleges to apply to
given the criteria described in this assignment. 

* Write one paragraph outlining the limitations of your 
analysis and what could be done in the future to address
those limitations.
